Could tides explain the extended stellar densities of the Sculptor and Ursa Minor dwarfs?

### 2025-12-08 paper/paper.tex
minor edits

  \abstract
  % context heading (optional)
   {Most dwarf spheroidals (dSph) of the Milky Way have light profiles following an exponential law, with a sharply declining outer density. However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
  % aims heading (mandatory)
   {We investigate whether tides may produce these extended light profiles.}
  % methods heading (mandatory)
   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way and Large Magellanic Cloud (LMC). }
  % results heading (mandatory)
   {Tides affect neither galaxy strongly. Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component evolves little. UMi instead loses about 95\% of its dark matter mass, but the stars would still appear unaffected with current observations. }
  % conclusions heading (optional), leave it empty if necessary
   {The outer stellar excess in Scl and UMi are likely not of tidal origin. Instead, these features are likely innate, possibly explained by past mergers, accretion, or star formation episodes. }

@@

Two radii provide estimates of where tidal effects occur.
The \textbf{Jacobi radius}, $r_{\rm J}$. approximates the radius where stars
become unbound in a galaxy\footnote{ 
    While strictly valid only for circular orbits, assuming
\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
lost during pericentric passages. $r_{\rm J}$ solves
}
\begin{equation}\label{eq:r_jacobi}
    3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
\end{equation}
given the mean enclosed  density of the host at pericentre $\bar \rho_{\rm MW}(r_{\rm peri})$,
and the mean enclosed satellite density $\bar \rho_{\rm dwarf}(r)$ \citep[eq. 7-84]{BT1987}.
If \(r_J\) is comparable to
the visible extent of a galaxy, we should expect to find clear signs of
tidal disturbance. 
The \textbf{break radius}, as defined in
\citet{penarrubia+2009}, marks the outermost dynamically-relaxed radius after pericentric passage. Based on the velocity dispersion, \(\sigma_v\), and time elapsed
since pericentre, \(\Delta t\), the break radius is
\begin{equation}\label{eq:r_break}
    r_{\rm break} = C\,\sigma_{v}\,\Delta t
\end{equation} 
where the empirical constant is \(C \approx 0.55\).

@@

Measuring the light profile of a resolved galaxy requires careful
consideration of whether any given star belongs to the system or not.
Without removing contamination from foreground/background sources, faint
features may be lost in the noise or be of uncertain association. In addition to colours and magnitudes, \emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallax and proper motion which improve membership assignment.

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data.\footnotemark{} J+24 used a Bayesian framework
incorporating proper motion, colour--magnitude diagram, and
spatial information to determine the probability that a given star
belongs to the satellite, $P_{\rm sat}$. By accounting for proper motions
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component, J+24 detected
candidate members out to ${\sim}10$ half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

@@

For our sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\).
%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with \(G < 21\).
Most stars have \(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact
choice of probability threshold has little effect on the resulting sample. Even
at the probability threshold of 0.2, the purity remains high when validated
against spectroscopic line-of-sight (LOS) velocities (${\sim}90\%$,
J+24).\footnotemark{} However, we find that our conclusions are unchanged when
limiting samples to only the brightest stars. Altogether, the J+24 method
provides a high-quality, low-contamination sample of dwarf galaxy candidate
member stars, which we will now investigate in further detail.

\footnotetext{This
  would indicate that the J+24 model probabilities are miss-calibrated.
  However, most LOS surveys of dwarf galaxies select brighter stars
  (which have better \emph{Gaia} measurements) and likely members,
  complicating the interpretation of this purity estimate.}

Fig.~\ref{fig:observed_density_profiles} compares the density
profiles of Scl, UMi, and eight other classical dwarf
galaxies.\footnote{
    We define classicals to be non-Magellanic dwarfs brighter than
    $Mv = -7.7$, i.e., Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II,
    Carina, Draco, Ursa Minor, Canes Venatici I, Sextans I, and Crater II. Of
    these dwarfs, we exclude Antlia II, due to the extremely high background,
    and Sagittarius, which was not included in J+24. 
}
All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions.\footnote{
    Comparing
    density profiles of dwarf galaxies is complicated by variations in the
    effective depth between galaxies. However, the deviations between Scl,
    UMi, and other galaxies are apparent even where the data is complete
    across all dwarfs.} 
In the outer regions, however, Scl and Ursa
Minor deviate and show a clear outer excess over an exponential law
(solid black line). These galaxies are better fit by a Plummer law
(dashed black line). The deviation from an exponential grows outwards,
and at \(\sim 8 R_h\), may reach 2 orders of magnitude.

@@

For our MW-only models, we adopt the potential from \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
The Potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
A NFW density profile is defined as,
\begin{equation}\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation}
where $r_s$ and $\rho_s$ are the scale mass and scale density. We later parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\). Equivalently, the dark matter halo has a virial mass\footnotemark{} \(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
\footnotetext{The virial mass is the mass enclosed within the region containing a mean density 200 times the Universe's critical density}
%The Galactic bulge is described by a \citet{hernquist1990} potential,
%\begin{equation}{
%    \Phi(r) = - \frac{GM}{r + a},
%}\end{equation} 
%where \(a=1.3\,{\rm kpc}\) is the scale radius and
%\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
%disks are represented with the \citet{miyamoto+nagai1975} cylindrical
%potential, 
%\begin{equation}{
%    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
%}\end{equation} 
%where \(a\) is the disc radial scale length, \(b\) is
%the scale height, and \(M\) is the total mass of the disk. For the thin
%disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
%\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
%\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).

@@

To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
smallest pericentre among all possible orbits integrated backwards for
\(10\,\Gyr\). We achieve this by selecting the median parameters of all
orbits with a pericentre less than the 0.0027th quantile pericentre,
yielding a pericentre of 43 kpc. Given the current observations, it is
unlikely that either has had a significantly smaller pericentre than
our selected orbit, which we refer to as the \smallperi{} orbit. We take
the first apocentre after a look-back time of 10 Gyr as the initial conditions for our models of Scl and UMi,
noted in Table~\ref{tbl:orbit_ics}.

For the Scl MW+LMC model, we select, instead, the $6\sigma$ smallest pericentre with respect to the LMC, (by doubling the observed uncertainties), and only integrate backwards for $2\,\Gyr$.

@@

We use \agama{} \citep{agama} to generate the initial N-body dark matter
halo. We assume galaxies are described by a spherical, isotropic 
dark matter halo (Eq.~\ref{eq:nfw}). The NFW
profile truncated in the outer regions by
\begin{equation}\label{eq:trunc_nfw}
    \rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
\end{equation}
where we adopt \(r_t = 20 r_s\).

@@

To ensure that the initial conditions of the simulation are dynamically
relaxed, we run a halo first in isolation using
\gadget{}. We use the same isolation run
for all halos (with $\rmax=6$ and $\vmax=31$, ran for $5\,\Gyr$) and rescale the results to the desired initial halo.
Our isolation halo is converged down to a radius $r\approx 100\,{\rm pc}$ (see also the convergence radius from \citealt{power+2003}).

@@

If we
know the distribution function\footnote{i.e., the phase-space density of
  particles. Note that since \(f\) is formally defined in 6D phase
  space, there is an additional `density of states' term in order to
  calculate the histogram of particles with \({\cal E}\),
  \(dM/d{\cal E}\) \citep[section 4.4.5,][]{BT1987}.} \(f({\cal E})\),
then we assign a stellar weight to a given particle with energy
\({\cal E}\) using \begin{equation}{
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
}\end{equation}
We use Eddington inversion to find the distribution
function, \begin{equation}{
    f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
}
\end{equation} 
where $\rho$ is the stellar or DM density, \(\Psi\) is the potential (normalized to vanish at infinity), and
\({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 \V^2\)  
\citep[see eq. 4-140b in][]{BT1987}. We
take \(\Psi\) and $\rho$ from the assumed analytic density profiles. We calculate \(\rho_\star\) from \(\Sigma_\star\) via the inverse Abel transform.

We consider both an exponential stellar profile, 
\begin{equation}\label{eq:exponential_law}
\Sigma_{\rm exp} = \Sigma_0\exp(-\alpha\,R / \alpha\,R_h).
\end{equation}
for central surface density $\Sigma_0$, half-light radius $R_h$, and $\alpha \approx 1.6783$.
We also consider a more extended Plummer law, 
\begin{equation}{
\Sigma_{\rm Pl} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
}\end{equation}
for half-light radius $R_h$.

Fig.~\ref{fig:initial_halos} shows the resulting initial conditions for the MW-only models before and after 5 Gyr of isolation evolution.

@@

The black lines in \ref{fig:orbits} show our selected orbits (black), representing an extreme low-pericentre orbit. We adopt these orbits as the most severe observationally-permisible tidal histories of each galaxy.

@@

A key caveat is that the long-term orbits of satellites are highly uncertain.
Analytic MW potentials neglect unknown details including triaxiality, mass
evolution, and substructure, causing analytic orbits to deviate significantly
from true satellite orbits \citep[e.g.,]{dsouza+bell2022, santistevan+2023}.
While we cannot constrain the long-term histories of these dwarfs, our results
indicate that {\it recent tidal effects} would not be observable in Scl and
UMi.

>>>>

  \abstract
  % context heading (optional)
   {Most dwarf spheroidals (dSph) of the Milky Way have light profiles following an exponential law, with a sharply declining outer density. However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
  % aims heading (mandatory)
   {We investigate whether tides may produce these extended light profiles.}
  % methods heading (mandatory)
   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way and Large Magellanic Cloud (LMC). }
  % results heading (mandatory)
   {Tides affect neither galaxy strongly. Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component evolves little. UMi instead loses about 95\% of its dark matter mass, but the stars should appear unaffected. }
  % conclusions heading (optional), leave it empty if necessary
   {The outer stellar excess in Scl and UMi are likely not of tidal origin. Instead, these features are likely innate, possibly explained by past mergers, accretion, or star formation episodes. }

@@

Two radii provide estimates of where tidal effects occur.
The \textbf{Jacobi radius}, $r_{\rm J}$, approximates the radius where stars
become unbound in a galaxy.\footnote{ 
    While strictly valid only for circular orbits, assuming
\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
lost during pericentric passages.} $r_{\rm J}$ solves
\begin{equation}\label{eq:r_jacobi}
    3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
\end{equation}
given the mean enclosed  density of the host at pericentre $\bar \rho_{\rm MW}(r_{\rm peri})$,
and the mean enclosed satellite density $\bar \rho_{\rm dwarf}(r)$ \citep[eq. 7-84]{BT1987}.
If \(r_J\) is comparable to
the visible extent of a galaxy, we should expect to find clear signs of
tidal disturbance. 
The \textbf{break radius}, as defined in
\citet{penarrubia+2009}, marks the outermost dynamically-relaxed radius after pericentric passage. Based on the velocity dispersion, \(\sigma_v\), and time elapsed
since pericentre, \(\Delta t\), the break radius is
\begin{equation}\label{eq:r_break}
    r_{\rm break} = C\,\sigma_{v}\,\Delta t
\end{equation} 
where the empirical constant is \(C \approx 0.55\).

@@

Measuring the light profile of a resolved galaxy requires careful
consideration of whether any given star belongs to the system or not.
Without removing contamination from foreground/background sources, faint
features may be lost in the noise or be of uncertain association. \emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallaxes and proper motions, improving membership by providing key additional information.

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion, colour--magnitude diagram, and
spatial information to determine the probability that a given star
belongs to the satellite, $P_{\rm sat}$. By accounting for proper motions
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component.\footnotemark{} J+24 detected
candidate members out to ${\sim}10$ half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

@@

For our sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\).
%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with \(G < 21\).
Most stars have \(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact
choice of probability threshold has little effect on the resulting sample. Even
at the probability threshold of 0.2, the purity remains high when validated
against spectroscopic line-of-sight (LOS) velocities (${\sim}90\%$,
J+24).\footnotemark{} Altogether, the J+24 method
provides a high-quality, low-contamination sample of dwarf galaxy candidate
member stars, which we will now investigate in further detail.

\footnotetext{This
  would indicate that the J+24 model probabilities are miss-calibrated.
  However, most LOS surveys of dwarf galaxies select brighter stars
  (which have better \emph{Gaia} measurements) and likely members,
  complicating the interpretation of this purity estimate. However, we find that our conclusions are unchanged when
limiting samples to only the brightest stars. }

Fig.~\ref{fig:observed_density_profiles} compares the density
profiles of Scl, UMi, and eight other classical dwarf
galaxies.\footnote{
    We define classicals to be non-Magellanic dwarfs brighter than
    $Mv = -7.7$, i.e., Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II,
    Carina, Draco, Ursa Minor, Canes Venatici I, Sextans I, and Crater II. Of
    these dwarfs, we exclude Antlia II, due to the extremely high background,
    and Sagittarius, which was not included in J+24. 
}
All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions.\footnote{
    Comparing
    density profiles of dwarf galaxies is complicated by variations in the
    effective depth between galaxies. However, the deviations between Scl,
    UMi, and other galaxies are apparent even where the data is complete
    across all dwarfs.} 
In the outer regions, however, Scl and Ursa
Minor deviate and show a clear outer excess over an exponential law
(solid black line). These galaxies are better fit by a Plummer law
(dashed black line). The deviation from an exponential grows outwards
and may reach 2 orders of magnitude at $8 R_h$.

@@

For our MW-only models, we adopt the potential from \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
The potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
A NFW density profile is defined as,
\begin{equation}\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation}
where $r_s$ and $\rho_s$ are the scale mass and scale density. We later parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\). Equivalently, the dark matter halo has a virial mass\footnotemark{} \(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
\footnotetext{The virial mass is the mass enclosed within the region containing a mean density 200 times the Universe's critical density}
%The Galactic bulge is described by a \citet{hernquist1990} potential,
%\begin{equation}{
%    \Phi(r) = - \frac{GM}{r + a},
%}\end{equation} 
%where \(a=1.3\,{\rm kpc}\) is the scale radius and
%\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
%disks are represented with the \citet{miyamoto+nagai1975} cylindrical
%potential, 
%\begin{equation}{
%    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
%}\end{equation} 
%where \(a\) is the disc radial scale length, \(b\) is
%the scale height, and \(M\) is the total mass of the disk. For the thin
%disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
%\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
%\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).

@@

To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
smallest pericentre among all possible orbits integrated backwards for
\(10\,\Gyr\). We achieve this by selecting the median parameters of all
orbits with a pericentre less than the 0.0027th quantile pericentre,
yielding a pericentre of 43 kpc. Given the current observations, it is
unlikely that either has had a significantly smaller pericentre than
our selected orbit, which we refer to as the \smallperi{} orbit. We take
the first apocentre after a look-back time of 10 Gyr as the initial conditions for our models of Scl and UMi,
noted in Table~\ref{tbl:orbit_ics}.
For the Scl MW+LMC model, we select, instead, the $6\sigma$ smallest pericentre with respect to the LMC, (by doubling the observed uncertainties). We only integrate this model backwards for $2\,\Gyr$.

@@

We use \agama{} \citep{agama} to generate the initial N-body dark matter
halo. We assume galaxies are described by a spherical, isotropic 
NFW dark matter halo (Eq.~\ref{eq:nfw}). The NFW
profile truncated in the outer regions by
\begin{equation}\label{eq:trunc_nfw}
    \rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
\end{equation}
where we adopt \(r_t = 20 r_s\).

@@

To ensure that the initial conditions of the simulation are dynamically
relaxed, we run a halo first in isolation. We use the same isolation run
for all halos (with $\rmax=6$ and $\vmax=31$, ran for $5\,\Gyr$) and rescale the results to the desired initial halo.
Our isolation halo is converged down to a radius $r\approx 100\,{\rm pc}$ (see also the convergence radius from \citealt{power+2003}).

@@

If we
know the distribution function\footnote{i.e., the phase-space density of
  particles. Note that since \(f\) is formally defined in 6D phase
  space, there is an additional `density of states' term in order to
  calculate the histogram of particles with \({\cal E}\),
  \(dM/d{\cal E}\) \citep[section 4.4.5,][]{BT1987}.} \(f({\cal E})\),
then we assign a stellar weight to a given particle with energy
\({\cal E}\) using \begin{equation}{
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
}\end{equation}
We use Eddington inversion to find the distribution
function, \begin{equation}{
    f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
}
\end{equation} 
where $\rho$ is the stellar or DM density, \(\Psi\) is the potential (normalized to vanish at infinity), and
\({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 \V^2\)  
\citep[see eq. 4-140b in][]{BT1987}. We
take \(\Psi\) and $\rho$ from the assumed analytic density profiles, calculating \(\rho_\star\) from \(\Sigma_\star\) via the inverse Abel transform.

We consider both an exponential stellar profile, 
\begin{equation}\label{eq:exponential_law}
\Sigma_{\rm exp} = \Sigma_0\exp(-\alpha\,R / \alpha\,R_h).
\end{equation}
for central surface density $\Sigma_0$, half-light radius $R_h$, and $\alpha \approx 1.6783$.
We also consider a more extended \citep{plummer1911} law component, 
\begin{equation}{
\Sigma_{\rm Pl} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
}\end{equation}
for half-light radius $R_h$. 
Fig.~\ref{fig:initial_halos} shows the resulting initial conditions for the MW-only models before and after 5 Gyr of isolation evolution.

@@

The black lines in \ref{fig:orbits} show our selected orbits (black), representing an extreme low-pericentre orbit. We adopt these orbits as the most severe observationally-permissible tidal histories of each galaxy.

@@

A key caveat is that the long-term orbits of satellites are highly uncertain.
Analytic MW potentials neglect unknown details including triaxiality, mass
evolution, and substructure, causing analytic orbits to deviate significantly
from true satellite orbits \citep[e.g.,][]{dsouza+bell2022, santistevan+2023}.
While we cannot constrain the long-term histories of these dwarfs, our results
indicate that {\it recent tidal effects} are likely not be observable in Scl and
UMi.
